Summary of tumor cell classification with `SingleR` for `r params$library_id`

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/03-copykat.Rmd

full_ck_result_paths <- c(
  no_ref = file.path(results_dir, "no_normal", full_ck_result_file),
  with_ref = file.path(results_dir, "endothelial_normal", full_ck_result_file)
)

```

@@

```{r}
# for every cell, calculate the mean detection level across all genes in a given chromosome
full_cnv_df <- full_ck_results_df |> 
  tidyr::pivot_longer(cols = -c(
    reference_used,
    chrom
    ),
    names_to = "barcodes",
    values_to = "cnv_detection") |> 
  dplyr::group_by(chrom, barcodes, reference_used) |> 
  dplyr::summarise(mean_cnv_detection = mean(cnv_detection))

@@

```{r, fig.height=15, fig.width=10}
# create faceted density plots showing estimation of CNV detection across each chr of interest 
# colored by aneuploid/diploid estimation 
ggplot(cnv_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
      geom_density() +
      theme_bw() +
      facet_grid(rows = vars(chrom), 
                 cols = vars(reference_used))
```

@@

caret_df_list |> 
  purrr::imap(\(df, ref_type){
    caret::confusionMatrix(
      table(
        df$marker_gene_classification, 
        df$copykat)
    )
  })
```

@@

```{r}
# calculate Jaccard similarity index for each reference type
jaccard_matrices <- caret_df_list |>
  purrr::map(\(df) {

    make_jaccard_matrix(
      df,
      "marker_gene_classification",
      "copykat.pred"
    )
  })
```

@@

# heatmaps comparing tumor/normal annotations manually vs. copyKAT
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(jaccard_mtx, ref_type) {
      ComplexHeatmap::Heatmap(
        t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = ref_type,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = ref_type == "no_ref",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

>>>>

full_ck_result_paths <- c(
  no_ref = file.path(results_dir, "no_normal", full_ck_result_file),
  with_ref = file.path(results_dir, "endothelial_normal", full_ck_result_file)
)
```

@@

```{r}
# for every cell, calculate the mean detection level across all genes in a given chromosome
full_cnv_df <- full_ck_results_df |>
  tidyr::pivot_longer(
    cols = -c(
      reference_used,
      chrom
    ),
    names_to = "barcodes",
    values_to = "cnv_detection"
  ) |>
  dplyr::group_by(chrom, barcodes, reference_used) |>
  dplyr::summarise(mean_cnv_detection = mean(cnv_detection))

@@

```{r, fig.height=15, fig.width=10}
# create faceted density plots showing estimation of CNV detection across each chr of interest
# colored by aneuploid/diploid estimation
ggplot(cnv_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
  geom_density() +
  theme_bw() +
  facet_grid(
    rows = vars(chrom),
    cols = vars(reference_used)
  )
```

@@

caret_df_list |>
  purrr::imap(\(df, ref_type){
    caret::confusionMatrix(
      table(
        df$marker_gene_classification,
        df$copykat
      )
    )
  })
```

@@

```{r}
# calculate Jaccard similarity index for each reference type
jaccard_matrices <- caret_df_list |>
  purrr::map(\(df) {
    make_jaccard_matrix(
      df,
      "marker_gene_classification",
      "copykat.pred"
    )
  })
```

@@

# heatmaps comparing tumor/normal annotations manually vs. copyKAT
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(jaccard_mtx, ref_type) {
      ComplexHeatmap::Heatmap(
        t(jaccard_mtx), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = ref_type,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = ref_type == "no_ref",
      )
    }
  ) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/cnv-workflow/01-marker-genes.Rmd

```{r}
# read in normal and tumor cells for later comparisons
if(file.exists(params$reference_cell_file)){
  ref_cells_df <- readr::read_tsv(params$reference_cell_file)
} else {
  ref_cells_df <- NULL
}
```

@@

```{r}
# test if _any_ marker genes have expression
if(sum(umap_df$gene_expression) == 0){
  has_marker_gene_exp <- FALSE
  message("No marker gene expression detected, so all plots of marker genes will be skipped.")
} else {
  has_marker_gene_exp <- TRUE
}
```

@@

# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

```{r, eval=has_marker_gene_exp}
# plot comparing marker gene expression in cells annotated as "normal" and "tumor" by SingleR and CellAssign
# use already identified normal/tumor cell types
if(!is.null(ref_cells_df)){

  # get list of normal cells
  normal_cells <- ref_cells_df |>
    dplyr::filter(reference_cell_class == "Normal") |>
    dplyr::pull(barcodes) |>
    unique()

@@

  # only create this plot if normal and tumor cells are provided
  if(length(normal_cells) > 0 && length(tumor_cells) > 0){

    # indicate cells that are part of normal and tumor lists
    select_celltype_df <- celltypes_df |>
      dplyr::mutate(
        automated_annotation = dplyr::case_when(
          barcodes %in% normal_cells ~ "Normal cell",
          barcodes %in% tumor_cells ~ "Tumor cell"
        )
      ) |>
      dplyr::select(barcodes, automated_annotation)

@@

```{r}
if(has_marker_gene_exp){
  # get a dataframe with barcodes, sum_classification and cluster_classification
  celltypes_df <- celltypes_df |>
    dplyr::select(
      barcodes,
      marker_gene_classification = sum_classification
    )

  readr::write_tsv(celltypes_df, output_classifications_file)
} else{
  # save marker gene classification as NA if no expression
  celltypes_df <- data.frame(
    barcodes = umap_df$barcodes,
    marker_gene_classification = NA
  )

>>>>

```{r}
# read in normal and tumor cells for later comparisons
if (file.exists(params$reference_cell_file)) {
  ref_cells_df <- readr::read_tsv(params$reference_cell_file)
} else {
  ref_cells_df <- NULL
}
```

@@

```{r}
# test if _any_ marker genes have expression
if (sum(umap_df$gene_expression) == 0) {
  has_marker_gene_exp <- FALSE
  message("No marker gene expression detected, so all plots of marker genes will be skipped.")
} else {
  has_marker_gene_exp <- TRUE
}
```

@@

# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }
  ) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

```{r, eval=has_marker_gene_exp}
# plot comparing marker gene expression in cells annotated as "normal" and "tumor" by SingleR and CellAssign
# use already identified normal/tumor cell types
if (!is.null(ref_cells_df)) {
  # get list of normal cells
  normal_cells <- ref_cells_df |>
    dplyr::filter(reference_cell_class == "Normal") |>
    dplyr::pull(barcodes) |>
    unique()

@@

  # only create this plot if normal and tumor cells are provided
  if (length(normal_cells) > 0 && length(tumor_cells) > 0) {
    # indicate cells that are part of normal and tumor lists
    select_celltype_df <- celltypes_df |>
      dplyr::mutate(
        automated_annotation = dplyr::case_when(
          barcodes %in% normal_cells ~ "Normal cell",
          barcodes %in% tumor_cells ~ "Tumor cell"
        )
      ) |>
      dplyr::select(barcodes, automated_annotation)

@@

```{r}
if (has_marker_gene_exp) {
  # get a dataframe with barcodes, sum_classification and cluster_classification
  celltypes_df <- celltypes_df |>
    dplyr::select(
      barcodes,
      marker_gene_classification = sum_classification
    )

  readr::write_tsv(celltypes_df, output_classifications_file)
} else {
  # save marker gene classification as NA if no expression
  celltypes_df <- data.frame(
    barcodes = umap_df$barcodes,
    marker_gene_classification = NA
  )

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/singler-workflow/01-singler-results.Rmd
Fix spelling errors

```{r}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(singler_results_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
  dplyr::mutate(
    # set factor order for aucell 
    aucell_annotation = forcats::fct_relevel(aucell_annotation, "tumor", after= 0), 
    # first remove other library IDs from tumor annotations
    # we only care about whether or not it was from the original library or a new library
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |> 
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later 
    singler_lumped = singler_annotation |>
          # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
          # this is why we ignore warnings in this chunk 
          forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf), 
    # make a label that is just tumor or normal
    singler_tumor_normal = dplyr::if_else(!(singler_annotation %in% c("tumor", "unknown")), "normal", singler_annotation) |> 
      forcats::fct_relevel("tumor", after = 0), 
    # get the consensus between aucell and singler 
    consensus = dplyr::case_when(
      singler_annotation == "tumor" & aucell_annotation == "tumor" ~ "tumor_both",
      singler_annotation == "tumor" & aucell_annotation != "tumor" ~ "tumor_singler",
      singler_annotation != "tumor" & aucell_annotation == "tumor" ~ "tumor_aucell",
      .default = "normal"
    ) |> 
      forcats::fct_relevel("normal", after = Inf)
  )
```

@@

```{r}
plot_density(classification_df,
             "tumor_sum",
             "singler_tumor_normal")
```

@@

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_tumor_normal")
  })
```

@@

```{r}
marker_gene_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_lumped")
  })
```

@@

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "singler_lumped")
  })
```

### Tumor cells annotatd by `SingleR` and `AUCell`

@@

```{r}
plot_density(classification_df,
             "tumor_sum",
             "consensus")
```

@@

```{r}
geneset_columns |> 
  purrr::map(\(column){
    plot_density(classification_df,
                 column,
                 "consensus")
  })
```

>>>>

```{r}
# generate classification df to use for plots
classification_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  # join with previous annotations, singler results, and gene set scores
  dplyr::left_join(singler_results_df, by = "barcodes") |>
  dplyr::left_join(geneset_scores_df, by = "barcodes") |>
  dplyr::mutate(
    # set factor order for aucell
    aucell_annotation = forcats::fct_relevel(aucell_annotation, "tumor", after = 0),
    # first remove other library IDs from tumor annotations
    # we only care about whether or not it was from the original library or a new library
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      # there are warnings because we have some that have less than 7 cell types (ones that only did tumor/normal)
      # this is why we ignore warnings in this chunk
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf),
    # make a label that is just tumor or normal
    singler_tumor_normal = dplyr::if_else(!(singler_annotation %in% c("tumor", "unknown")), "normal", singler_annotation) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the consensus between aucell and singler
    consensus = dplyr::case_when(
      singler_annotation == "tumor" & aucell_annotation == "tumor" ~ "tumor_both",
      singler_annotation == "tumor" & aucell_annotation != "tumor" ~ "tumor_singler",
      singler_annotation != "tumor" & aucell_annotation == "tumor" ~ "tumor_aucell",
      .default = "normal"
    ) |>
      forcats::fct_relevel("normal", after = Inf)
  )
```

@@

```{r}
plot_density(
  classification_df,
  "tumor_sum",
  "singler_tumor_normal"
)
```

@@

```{r, fig.height=10}
geneset_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "singler_tumor_normal"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```

@@

```{r, fig.height=10}
marker_gene_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "singler_lumped"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```

@@

```{r, fig.height=10}
geneset_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "singler_lumped"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```

### Tumor cells annotated by `SingleR` and `AUCell`

@@

```{r}
plot_density(
  classification_df,
  "tumor_sum",
  "consensus"
)
```

@@

```{r, fig.height=10}
geneset_columns |>
  purrr::map(\(column){
    plot_density(
      classification_df,
      column,
      "consensus"
    )
  }) |>
  patchwork::wrap_plots(ncol = 1)
```
